---
title: Setting the random seed in Bioconductor packages
author: 
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
package: BiocSeed
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{BioC'ing the seed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Overview

When using functions that have a stochastic component, it is standard practice to manually call `set.seed()` to obtain reproducible results.
However, this is easy to forget during interactive analyses and is inconvenient when writing scripts.
For example, if the seed is not set prior to every function call involving randomization,
the reproducibility of any non-trivial workflow becomes highly fragile as changes to the PRNG stream will affect all subsequent steps.

The general solution is for developers to set the seed on behalf of the user within each function that implements random procedures.
This provides robust reproducibility without requiring the user to possess deep knowledge of each function's implementation, but leaves us with the question of "which seed?"
Using a constant seed for all function calls exposes us to potential biases as the same stream of pseudo-random numbers would be used everywhere;
this may compromise the long-run correctness of particular statistical methods.
For example, if a particular seed resulted in a stretch of non-extreme test statistics for Monte Carlo hypothesis tests, we would become consistently anticonservative for all datasets.

To this end, `r Biocpkg("BiocSeed")` provides the `setBiocSeed()` function to set the seed to a "deterministically variable" value within Bioconductor packages.
The central idea is that the seed is itself a function of the data.
This ensures that the same call on the same data will reproducibly yield the same result, while different calls will use different seeds.
In this manner, we avoid the statistical problems associated with using the same seed in all situations.

# Basic usage

Given an integer or character vector or a list thereof, `setBiocSeed()` will compute a 32-bit hash and call `set.seed()` on the resulting value.
A typical function definition might look like this:

```{r}
library(BiocSeed)
my_somewhat_random_task <- function(x) {
    info <- setBiocSeed(head(x))
    on.exit(unsetBiocSeed(info))

    # Do somewhat random things here.
    sample(x)
}
```

This ensures that identical calls on the same data yield the same results:

```{r}
my_somewhat_random_task(letters)
my_somewhat_random_task(letters)
```

If we change the input data, we get different (but still reproducible) results:

```{r}
my_somewhat_random_task(LETTERS)
my_somewhat_random_task(LETTERS)
```

# Generating the seed

We limit `setBiocSeed()`'s inputs to integer and character values, so as to avoid variation in the seed across machines due to differences in numerical precision.
Directly supplying numeric inputs will raise an error:

```{r}
try(setBiocSeed(0.01))
```

Instead, developers should supply other attributes of numeric inputs, most typically the length or dimensions.
This should provide a decent amount of variation in genomics applications with differences in the number of samples/features across studies. 

```{r}
my_somewhat_random_task2 <- function(x) {
    # Assuming 'x' is a matrix.
    info <- setBiocSeed(dim(x))
    on.exit(unsetBiocSeed(info))

    # Do somewhat random things here.
    sample(x)
}

# Different results:
my_somewhat_random_task2(rbind(1:10))
my_somewhat_random_task2(matrix(1:10, 5, 2))
```

Even better is to include a unique identifier in the `setBiocSeed()` call within each function.
This ensures that different functions will use different seeds, even if they receive the same length/dimensions of their inputs.
Some readers may recognize the similarity with the concept of "salting" from crytography.

```{r}
my_somewhat_random_task3a <- function(x) {
    # Assuming 'x' is a matrix.
    info <- setBiocSeed(list("this is 3a", dim(x)))
    on.exit(unsetBiocSeed(info))

    # Do somewhat random things here.
    sample(x)
}

my_somewhat_random_task3b <- function(x) {
    # Assuming 'x' is a matrix.
    info <- setBiocSeed(list("this is 3b", dim(x)))
    on.exit(unsetBiocSeed(info))

    # Do somewhat random things here.
    sample(x)
}

# Reproducible within functions but different across functions.
my_somewhat_random_task3a(rbind(1:5))
my_somewhat_random_task3a(rbind(1:5))
my_somewhat_random_task3b(rbind(1:5))
my_somewhat_random_task3b(rbind(1:5))
```

It is also possible to include other (integer or character) arguments in the list supplied to `setBiocSeed()`, which adds further diversity to the choice of seed.
However, this should be done carefully to avoid surprising changes to the results.
To illustrate, the `bad_example` function below provides an `n=` argument to take the `head()` of the sampled values within the function.
A user might reasonably expect that they could get the same results by calling `bad_example()` and take the `head()` outside the function, 
but this would not be the case if `n` were included in the `setBiocSeed()` inputs.

```{r}
bad_example <- function(x, n=length(x)) {
    # Assuming 'x' is a matrix.
    info <- setBiocSeed(list("this is good", length(x), n))
    on.exit(unsetBiocSeed(info))

    head(sample(x), n=n)
}

bad_example(1:10, n=5L)
head(bad_example(1:10), 5)
```

As a general rule, developers should only include other arguments in `setBiocSeed()` if those arguments change the results by altering the PRNG state.
In such cases, there is no reason for the user to expect that they could recover the same results outside the function if they used different values for those arguments.

# Respecting the global seed

`unsetBiocSeed()` will restore the global seed to its previous state when the function exits.
This behavior avoids surprising changes to the user-level results when functions silently advance the random number stream. 
For example, end-users can expect to do things like:

```{r}
set.seed(1000)
runif(1)

set.seed(1000)
out <- my_somewhat_random_task(letters)
runif(1) # same result as before
```

This is also helpful in applications involving iterative simulations where each iteration calls our `r Biocpkg("BiocSeed")`-containing function.
Failing to restore the global seed would cause all iterations to be identical if the seed is always reset to some constant value.

```{r}
set.seed(1000)

niters <- 10
collected <- numeric(niters)
for (i in seq_len(niters)) {
    out <- my_somewhat_random_task(letters)
    collected[i] <- runif(1)
}

collected # all values are different
```

We suggest wrapping `unsetBiocSeed()` in an `on.exit()` statement to ensure that the global seed is restored upon both function return or error.

# Turning it all off

If users want to respond to the global seed only, the `r Biocpkg("BiocSeed")` system can be disabled with `disableBiocSeed()`.

```{r}
set.seed(10)
disableBiocSeed()
my_somewhat_random_task(letters)
my_somewhat_random_task(letters)
```

It is also straightforward to re-enable it:

```{r}
enableBiocSeed()
my_somewhat_random_task(letters)
my_somewhat_random_task(letters)
```

Alternatively, developers may wish to pass along `seed=` argument to `setBiocSeed()`.
With `seed=NULL`, this will derive a seed from the input arguments, otherwise it will conveniently enable the user to set their own seed.
This mode may be useful if the seed derived from `x` is somehow inappropriate, or for testing out the effect of stochasticity on the function outputs.

```{r}
FUN <- function(x, seed=NULL) {
    info <- setBiocSeed(head(x), seed=seed)
    on.exit(unsetBiocSeed(info))
    sample(x)
}

FUN(1:10)
FUN(1:10)
FUN(1:10, seed=100)
```

# Session information {-}

```{r}
sessionInfo()
```
