---
title: Setting the random seed in Bioconductor packages
author: 
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
package: BiocSeed
output:
  BiocStyle::html_document:
    toc_float: yes
vignette: >
  %\VignetteIndexEntry{BioC'ing the seed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Overview

In some situations, an algorithm is theoretically deterministic but the implementation has a stochastic component, e.g., for initialization.
It would be helpful if we could abstract away the stochastic component so that users do not have to manually call `set.seed()` to obtain reproducible results.
To this end, `r Biocpkg("BiocSeed")` provides the `setBiocSeed()` function to set the seed to a "deterministically variable" value within Bioconductor packages.

The central idea is that the seed is itself a function of the data.
This ensures that the same call on the same data will reproducibly yield the same result, while different calls will use different seeds.
In this manner, we avoid using the same seed in all situations, which could lead to problems due to, e.g., statistical biases or poor initializations.

# Basic usage

`r Biocpkg("BiocSeed")` is intended for package developers who are writing functions with some stochastic component.
It can be used by passing some relevant slice of the function's data to `setBiocSeed()` as shown below.

```{r}
library(BiocSeed)
my_somewhat_random_task <- function(x) {
    info <- setBiocSeed(head(x))
    on.exit(unsetBiocSeed(info))

    # Do somewhat random things here.
    sample(x)
}
```

This ensures that identical calls on the same data yield the same results:

```{r}
my_somewhat_random_task(letters)
my_somewhat_random_task(letters)
```

If we change the input data, we get different (but still reproducible) results:

```{r}
my_somewhat_random_task(LETTERS)
my_somewhat_random_task(LETTERS)
```

# Respecting the global seed

`unsetBiocSeed()` will restore the global seed to its previous state when the function exits.
This behavior avoids surprising changes to the user-level results when functions silently advance the random number stream. 
For example, end-users can expect to do things like:

```{r}
set.seed(1000)
runif(1)

set.seed(1000)
out <- my_somewhat_random_task(letters)
runif(1) # same result as before
```

We suggest wrapping `unsetBiocSeed()` in an `on.exit()` statement to ensure that the global seed is restored upon error.

# Nested calls 

Developers are free to nest these calls, provided that `unsetBiocSeed()` is always called at each level.

```{r}
OUTER <- function(x) {
    info <- setBiocSeed(head(x))
    on.exit(unsetBiocSeed(info))

    INNER(x)
}

INNER <- function(x) {
    info <- setBiocSeed(head(x))
    on.exit(unsetBiocSeed(info))
    sample(x)
}

OUTER(1:10)
```

Note that the `setBiocSeed()` invocation in `INNER()` has no effect once the same call is made in `OUTER()`.
In a nested stack of `setBiocSeed()` calls, the seed is only set once on the first call.
This is intentional and allows the PRNG to progress through the random number stream,
avoiding any biases from limiting the stream to the sequence around the $2^{32}-1$ seed locations.

# Turning it all off

If users want to respond to the global seed only, the `r Biocpkg("BiocSeed")` system can be disabled with `disableBiocSeed()`.

```{r}
set.seed(10)
disableBiocSeed()
my_somewhat_random_task(letters)
my_somewhat_random_task(letters)
```

It is also straightforward to re-enable it:

```{r}
enableBiocSeed()
my_somewhat_random_task(letters)
my_somewhat_random_task(letters)
```

# Session information {-}

```{r}
sessionInfo()
```
